# Analyzing wines physicochemical properties by Marcos Alves

***


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(tidyr)
library(purrr)
library(gridExtra)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
white <- read.csv("winequality-white.csv", sep = ";")
red <- read.csv("winequality-red.csv", sep = ";")
```
# Introduction

In this project, I'm trying to identify what are the characteristics of good wine.

I'm using the public dataset **Wine Quality Datasets** - P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.
It contains two datasets, one for red wines and another for white wines.

  [Here](http://www3.dsi.uminho.pt/pcortez/wine/) you can find more information and the files used in this project.

The dataset uses the measurements of 13 chemical constituents from [Vinho Verde Wines](http://www.vinhoverde.pt/), a wine brand from the northwest region of Portugal.

## Variables
- Fixed acidity: g(tartaric acid)/dm3
- Volatile acidity: g(acetic acid)/dm3
- Citric acid: g/dm3
- Residual sugar: g/dm3
- Chlorides: g(sodium chloride)/dm3 the amount of salt in the wine
- Free sulfur dioxide: mg/dm3
- Total sulfur dioxide: mg/dm3
- Density: g/cm3
- pH
- Sulphates:g(potassium sulphate)/dm3
- Alcohol: % vol.
- Quality:  evaluated by a minimum of three sensory assessors (using blind tastes), classified in a scale from 0 (very bad) to 10 (excellent). The final sensory score is the median of these evaluations.

# Considerations

Wine certification assessed by sensory tests is very subjective, so it might be hard to find a clear correlation of the variables with the quality of the wine.
This is an interesting [Vox video](https://www.youtube.com/watch?v=mVKuCbjFfIY) about wines taste and price.

For this project, I will use the white wine (winequality-white.csv) dataset.

# Data Analysis - White Wine Dataset

***

# Univariate


```{r echo=FALSE, summary}

summary(white)

```

Dataset summary

```{r echo=FALSE, Structure}

str(white)

```

This is the structure of the dataset

```{r echo=FALSE, Head}

head(white)

```

These are the first 6 rows of each variable

```{r echo=FALSE, correlation_quality}

cor(white[,2:12], white$quality)

```

The correlation between each variable and quality 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_quality}

ggplot(aes(x = quality), data = white) +
  geom_bar(fill = "#8989ff") +
  geom_vline(xintercept = median(white$quality), color = "red")

```

Most wines have a quality of 6 and the best variable for quality correlation is alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE, hist_alcohol}

ggplot(aes(x = alcohol), data = white) +
  geom_bar(fill = "#1ab385", binwidth = .05) +
  geom_vline(xintercept = median(white$alcohol), color = "orange")

```

The histogram of alcohol

***

There are 4898 observations for this dataset (not so large) and all variables are numbers, except for the quality variable, which is an integer. Later I'll transform the quality variable to factor, so it can be treated as a label. 

For this analysis I'm not considering the outliers, so I can make the analysis more robust.

The residual sugar has some unusual distribution, where there is a high peak below 5 g/dm3, and also the alcohol seems to have a more low variance distribution. 

For the quality distribution, the majority of values are between 5 and 7, and very small quantities for the remaining score values, making it a low variance distribution. This is an indication that most of these brand wines have a good score.

The correlation of the variables with quality shows that these variables don't have a high value. The highest value is 0.44 for alcohol and then the inverse correlation with density (-0.31). These variables might be the best for the analysis.

# Bivariate

#### Let's create a new column with the quality values as labels so we can create a boxplot by quality.

```
white$scale <- as.factor(white$quality)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol}

white$scale <- as.factor(white$quality)

ggplot(aes(factor(quality), 
            alcohol), 
        data = white) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Quality")

```

> There is a clear indication of the correlation of alcohol and quality above the score of 6 


### Alcohol x Quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol2}

ggplot(aes(x = scale, y = alcohol), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlab("Wine Quality") +
    ylab("Alcohol")

```

### Alcohol and Quality Correlation:

```{r echo=FALSE}
cor(white$alcohol, white$quality)

```

## ggpairs

using the ggpair function we can have a look on all the relations between the variables

```{r echo=FALSE,  message=FALSE, warning=FALSE, fig.width=10, fig.height=10, ggpairs}
#space

set.seed(1990)
ggpairs(white[sample.int(nrow(white), 1000), ], upper = list(continuous = wrap("cor", size = 4))) +
  theme(axis.text = element_blank())##TAKES SOME TIME DO RUN

ggcorr(white)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, bi_Alcohol_x_density}

ggplot(aes(x = density, y = alcohol), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlim(0.987, 1.004) +
    xlab("Density") +
    ylab("Alcohol")

```

This plot has a good correlation, where the data is well distributed along the smooth line, and has a good inverse slope.

``` {r warning=FALSE, cor_Alcohol_x_density}

cor(white$alcohol, white$density)

```

### Citric Acid x Quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, citric.acid}

ggplot(aes(factor(quality), 
            citric.acid), 
        data = white) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'green')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "blue", 
               shape = 8, 
               size = 4) +
  xlab("Quality") +
  ylim(0, 1)

```

Most values of Citric Acid are in the same range, but in the score of 9 it seems to be a bit higher.

### Free sulfur dioxide x Quality:
```{r echo=FALSE, message=FALSE, warning=FALSE, freesulfurdioxide}

ggplot(aes(factor(quality), 
            free.sulfur.dioxide), 
        data = white) +
  geom_jitter( alpha = .3, color = "orange")  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Quality") +
  ylim(0,150)

```

Values above the quality of four are close to 35. There are a few outliers with hight values of free SO2 for the quality of 3, so I ajusted the y axis to have a better view where most of the values are. It also seems that the value of free SO2 has a small decrease above the quality of 8.

### pH x Quality:
```{r echo=FALSE, message=FALSE, warning=FALSE, pH}

ggplot(data=white, aes(x = scale, y = pH)) +
  geom_jitter( alpha = .3, color = "orange")  +
  geom_boxplot( alpha = .5,color = 'blue') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Quality")

```

There is a slight increase in pH as quality gets higher

pH seems to have some correlation with quality, so let's see what variables might affect it:

```{r echo=FALSE, message=FALSE, warning=FALSE, correlation_pH}

cor(white[,2:12], white$pH)

```

All the correlations seem low.

### residual.sugar x pH:

```{r echo=FALSE, message=FALSE, warning=FALSE, residual.sugar_x_pH}

ggplot(aes(x = residual.sugar, y = pH), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlim(0, 25) +
    xlab("residual sugar") +
    ylab("pH")

```

Here we can see the high peak from the Univariate analysis for the residual.sugar variable. This correlation looks weak.

### citric.acid x pH:

```{r echo=FALSE, message=FALSE, warning=FALSE, citric.acid_x_pH}

ggplot(aes(x = citric.acid, y = pH), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlab("Citric Acid") +
    ylab("pH")

```

In this plot most values are around 0.3 in the Citric Acid axis and a few pontual values in the 0.5 and 0.75, creating the shape of a line.

### sulphates x pH:

```{r echo=FALSE, message=FALSE, warning=FALSE, sulphates_x_pH}

ggplot(aes(x = sulphates, y = pH), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlab("Sulphates") +
    ylab("pH")

```

In this plot the distribution of data looks a bit better. These three plots of the relation of pH with other variables confirms that most correlations of pH are low. Maybe using another variable we'll be able to find some pattern. 

### Volatile Acidity x Quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, free.sulfur.dioxide}

ggplot(data=white, aes(x = scale, y = volatile.acidity)) +
  geom_jitter( alpha = .3, color = "orange")  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  xlab("Quality")


```

For Volatile Acidity, the values are below 30 for values of quality above 4, and the means look constant.

### Chlorides x Quality:

```{r echo=FALSE, message=FALSE, warning=FALSE, chlorides}

ggplot(data=white, aes(x = scale, y = chlorides)) +
  geom_jitter( alpha = .3, color = "orange")  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) +
  ylim(0, 0.1) +
  xlab("Quality")

```

This plot has a small range of values for Chlorides, so I ajusted the y axis to have a better view. With this change, it is easier to see a decline in Chlorides as the quality increases. This corroborates with the early analysis of correlations, where this plot has a value of -0.21.

### Residual Sugar x Chlorides

```{r echo=FALSE, message=FALSE, warning=FALSE, residual.sugar_x_chlorides}

ggplot(aes(x = chlorides, y = residual.sugar), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    xlab("Chlorides") +
    ylab("Residual Sugar") +
  ylim(0, 30)
  

```

Here, the residual.sugar doesn't look to be affected by the Chlorides, where most of the data makes almost a vertical shape.

***

Using the Bivariate plots, the strongest relationship found was the alcohol x density (-0.78). High quality values tend to have a high % of alcohol and lower density. These variables might be the most influencial for the tests of quality.

There is an unusual pattern, where some values of very low quality wines are almost the same of good quality wines, as seen in volatile.acidity x Quality; pH x Quality; free.sulfur.dioxide; citric.acid x Quality.

Using a third variable in the plot might help seeing better the relation of these variables with quality.

# Multivariate

### Density x Alcohol

```{r echo=FALSE,  message=FALSE, warning=FALSE, alcohol_x_density}

ggplot(aes(x = alcohol, y = density, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  ylim(0.987,1.003) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('Density by Alcohol and Quality')

```

This plot shows the inverse correlation between density and alcohol and a good color distinguish where most of the best wines are above 11% alcohol and 0.993 density. The regression line for the wines with quality iquals nine, looks to have a strong correlation for these variables, close to one.


## Density x Residual Sugar

```{r echo=FALSE,  message=FALSE, warning=FALSE, residual.sugar_x_density}

ggplot(aes(x = residual.sugar, y = density, color = scale), data = white) +
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  xlim(0, 20) +  ylim(0.987,1.002) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('Density by residual.sugar and Quality')

```

In the second Multivariate plot, we can see the density and residual.sugar correlation. There is a nice distinction of colors, separated by the smooth line. Most higher quality wines are below the line, thus we might argue that testers prefer a higher residual.sugar/density ratio. In this plot the score of nine also has a better corellation for the variables (looks close to 1).

###  Sugar x pH

```{r echo=FALSE,  message=FALSE, warning=FALSE, sugar_pH_quality}

ggplot(aes(x = residual.sugar, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  xlim(0, 20) +
  ggtitle('pH by Residual Sugar and Quality')

```

This plot does not add much value for the analysis. The regression lines don't have such a good slope and the colors are randomly distributed, with no well defined pattern, though the regression line for nine is separate from the rest.

###  Citric Acid and pH

```{r echo=FALSE,  message=FALSE, warning=FALSE, citric.acid_pH_quality}

ggplot(aes(x = citric.acid, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('pH by Citric Acid and Quality')
```

Here we can see that there is a interval in the Citric Acid Above 0.25 and below 0.5 where the wine quality is considerably higher and the pH doesn't interfere so much. The reggresion lines show some correlation for the score of eight, inverse correlation for seven and 4, and good inverse correlation for scores of 3 . This might indicate that low values of citric acid are found in low quality wines.

## Total Sulfur Dioxide x Free Sulfur Dioxide

```{r echo=FALSE,  message=FALSE, warning=FALSE, totalSO2_x_freeSO2}

ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  xlim(0, 100) +
  ylim(0, 280) +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('Total.sulfur.dioxide by free.sulfur.dioxide and Quality')

```

Some correlation is notible between these two variables. We can distinguish an area where there is a concentration of most green dots, in the lower part of the scatter points, wich indicates a preference for a higher free.sulfur.dioxide/total.sulfur.dioxide ratio. Maybe there is some stage in the process of winemaking that directs affects this. There is a interesting information about the regression line of scores of nine, where it is almost horizontal. Maybe really good quality wines don't get affected by this correlation.

### pH x Alcohol

```{r echo=FALSE,  message=FALSE, warning=FALSE, alcohol_x_pH}

ggplot(aes(x = alcohol, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
 # ylim(0.987,1.003)
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('pH by Alcohol and Quality')

```

In this plot, the correlation between pH and alcohol is very low, and the quality is guided mainly by the alcohol quantity. Again we can notice that quality values iqual nine have some correlation with these variables.

### pH x Volatile Acidity

```{r echo=FALSE,  message=FALSE, warning=FALSE, residualxpH}

ggplot(aes(x = volatile.acidity, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('pH by Volatile Acidity and Quality')

```

In this plot I was surprise by this noticeable distinction. For the extreme values of quality (3 and 9), these variables have an oposite regression line. The best quality wines have a correlation of pH and Volatile Acidity positive and the lowest quality wines have an inverse correlation, and the other values don't have expressive correlations. Also there are some color region distinction for the data points.

### pH x sulphates

```{r echo=FALSE,  message=FALSE, warning=FALSE, sulphatesxpH}

ggplot(aes(x = sulphates, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('pH by Sulphates and Quality')

```

In this plot the correlation between the variables is low but the regression line for nine separates from the rest and there is the oposite correlation for the extreme values of quality.

# Final Plots and Summary

## Plot One
```{r echo=FALSE, fig.width=10, fig.height=10, Plot_One}

ggplot(aes(x = scale, y = alcohol), data = white) +
    geom_point(position = 'jitter', color = "#46003d", alpha = 1/5) +
    geom_smooth(method='lm', aes(group = 1), color = "cyan") +
    ggtitle('Wine Quality by Alcohol Content') +
    xlab("Wine Quality") +
    ylab("Alcohol [%]")

```

### This was my first plot because of the quick alanysis of the correlation between the variables and quality. Alcohol got a score well above the others.  I think it is a good plot because it shows a direct influence of a substance in the quality of white wines.

## Plot Two
```{r echo=FALSE, warning=FALSE, fig.width=10, fig.height=10, Plot_Two}

ggplot(aes(x = residual.sugar, y = density, color = scale), data = white) +
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'div', guide = guide_legend(title = 'Quality')) +
  xlim(0, 20) +  ylim(0.987,1.002) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('Density by Residual Sugar and Quality') +
  xlab("Residual Sugar [g/dm3]") +
  ylab("Density [g/cm3]")
  

```

### For me this plot has the best correlation between the variables and it shows clearly a separation of good and bad wines. Also the regression line for the quality value of 9 indicates a very good correlation between Density and Residual sugar. I would strongly recommend to use this plot for selecting a good white wine (with a high value of alcohol, of course).

## Plot Three
```{r echo=FALSE, fig.width=10, fig.height=10, Plot_Three}

ggplot(aes(x = volatile.acidity, y = pH, color = scale), data = white) + 
  geom_point(position = 'jitter') +
  scale_color_brewer(type = 'seq', guide = guide_legend(title = 'Quality')) +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  ggtitle('pH by Volatile Acidity and Quality') +
  xlab("Volatile Acidity [g(tartaric acid)/dm3]") +
  ylab("pH")

```

### This plot was the one that I've got the most surprised. It was the first one the I notice a clear oposite correlation for the extreme values of quality. Even though this plot gives some good insight, I wouldn't use it for making decisions because the distribution of quality is very low for this values and the bulk of the data is in the 5-7 range.

************************

# Reflection

These are only the most noticeable physical features that could influence the quality of white wines found on this dataset. There are many other variables that could have an influence on the quality of a wine. I used the ggpairs function to help me find the most relevant features.

Even though the dataset is not so large and wine quality classification is very subjective, there are some correlations between the physicochemical properties that we can notice. The most noticeable is the alcohol %. Higher quality wines have a higher percentage of alcohol, which is correlated with lower density and a lower residual sugar proportion. All of these features must have a considerable impact on the color, smell, and taste of the wine.

These are values range I would pick for selecting a good whine:

- Fixed acidity: 6-7 g(tartaric acid)/dm3
- Volatile acidity: 0.20-0.35 g(acetic acid)/dm3
- Citric acid: 0.25-0.50 g/dm3
- Residual sugar: 1-6 g/dm3
- Chlorides: 0.03-0.04 g(sodium chloride)/dm3
- Free sulfur dioxide: 30-50 mg/dm3
- Total sulfur dioxide: 100-120 mg/dm3
- Density: 0.988-0.991 g/cm3
- pH: 3.2-3.3
- Sulphates:0.4-0.8 g(potassium sulphate)/dm3
- Alcohol: above 11 % vol.

In addition to this project, it would be interesting to use some machine learning algorithm to analyze this dataset and also analyze the red wine dataset for comparison.
